sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-03 doc/paper-jss/owls.Rmd
draft pcod captions

```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="Mesh used for model is based on X and Y that use an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The facet colours represent the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="This mesh uses an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The colour of each year represents the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

### 2022-11-03 doc/paper-jss/pcod.Rmd
draft pcod captions

For example, a mesh with a cutoff distance of 10 km can be made:

@@

```{r plot-mesh, cache=TRUE, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE}
plot(mesh)
```

@@

We can inspect simulated residuals from the DHARMa package:

```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE}
set.seed(123)
simulate(fit, nsim = 500) %>%
  dharma_residuals(fit)
```

Or perhaps better, use MCMC residuals:

@@

```{r mcmc-resids-vis, out.width="4in"}
qqnorm(r)
qqline(r)
```

@@

There are four kinds of predictions that we get out of the model. First we will show the predictions that incorporate all fixed effects and random effects:

```{r plot-all-effects}
plot_map(pred_qcs, "exp(est)") +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("Prediction (fixed effects + all random effects)")
```

We can also look at just the fixed effects, here `year`:

```{r plot-fix-defects}
plot_map(pred_qcs, "exp(est_non_rf)") +
  ggtitle("Prediction (fixed effects only)") +
  scale_fill_viridis_c(trans = "sqrt")
```

We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

```{r plot-spatial-effects}
plot_map(pred_qcs, "omega_s") +
  ggtitle("Spatial random effects only") +
  scale_fill_gradient2()
```

Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations. These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model.

```{r plot-spatiotemporal-effects}
plot_map(pred_qcs, "epsilon_st") +
  ggtitle("Spatiotemporal random effects only") +
  scale_fill_gradient2()
```

@@

This returns a matrix with a row for each row of the `newdata` data frame and a column for each replicate draw. We can use this matrix to calculate the standard deviation or coefficient of variation:

```{r sims-cv-plot, cache=TRUE}
qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
ggplot(qcs_grid, aes(X, Y, fill = cv)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10", option = "D")
```

We can calculate a standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE`. This returns the \pkg{TMB} objective function, which is required to pass the output to `get_index()`.

```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE}
qcs_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = qcs_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = qcs_grid$area, bias_correct = TRUE)

>>>>

For example, a mesh with a cutoff distance of 10 km can be made (Fig.\ \ref{fig:mesh}).

@@

```{r plot-mesh, cache=TRUE, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="This mesh uses a cutoff distance of 10 km.", fig.pos='h', fig.align='center'}
plot(mesh)
```

@@

We can inspect simulated residuals from the DHARMa package (Fig.\ \ref{fig:sim-resids}).

```{r sim-resids, out.width="4in", warning=FALSE, results='hide', cache=TRUE, echo=TRUE, fig.cap="Simulated residuals from the DHARMa package.", fig.pos='h', fig.align='center'}
set.seed(123)
simulate(fit, nsim = 500) %>%
  dharma_residuals(fit)
```

Or perhaps better, use MCMC residuals (Fig.\ \ref{fig:mcmc-resids}).

@@

```{r mcmc-resids-vis, out.width="4in", fig.cap="MCMC residuals.", fig.pos='h', fig.align='center'}
qqnorm(r)
qqline(r)
```

@@

There are four kinds of predictions that we get out of the model. First we will show the predictions that incorporate all fixed effects and random effects (Fig.\ \ref{fig:plot-all-effects}).

```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='h', fig.align='center'}
plot_map(pred_qcs, "exp(est)") +
  # ggtitle("Prediction (fixed effects + all random effects)") +
  scale_fill_viridis_c(trans = "log10") 
```

We can also look at just the fixed effects, here `year` (Fig.\ \ref{fig:plot-fix-defects}).

```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='h', fig.align='center'}
plot_map(pred_qcs, "exp(est_non_rf)") +
  # ggtitle("Prediction (fixed effects only)") +
  scale_fill_viridis_c(trans = "sqrt")
```

We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects}). In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='h', fig.align='center'}
plot_map(pred_qcs, "omega_s") +
  # ggtitle("Spatial random effects only") +
  scale_fill_gradient2()
```

Finally, we can look at the spatiotemporal random effects that represent deviations from the fixed effect predictions and the spatial random effect deviations (Fig.\ \ref{fig:plot-spatiotemporal-effects}). These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model.

```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='h', fig.align='center'}
plot_map(pred_qcs, "epsilon_st") +
  # ggtitle("Spatiotemporal random effects only") +
  scale_fill_gradient2()
```

@@

This returns a matrix with a row for each row of the `newdata` data frame and a column for each replicate draw. We can use this matrix to calculate the standard deviation or coefficient of variation (Fig.\ \ref{fig:sims-cv-plot}).

```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='h', fig.align='center'}
qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
ggplot(qcs_grid, aes(X, Y, fill = cv)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10", option = "D")
```

We can calculate a standardized population index by predicting on our survey domain grid and setting `return_tmb_object = TRUE` (Fig.\ \ref{fig:pcod-app-index}). This returns the \pkg{TMB} objective function, which is required to pass the output to `get_index()`.

```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='h', fig.align='center'}
qcs_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = qcs_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = qcs_grid$area, bias_correct = TRUE)
